Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-07-03 doc/MS_v4.Rmd
updates DOIs to ref. Moved Tables around

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ). Mean MLH1 focus counts for `r sum(main.table$Nmice)` mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes). **insert statement for cell sample size number**

@@

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; p < `r 1/1000000` -- 10-6). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval`). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

@@

We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 1). Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus*, *M. spretus* and _M. caroli_. We subsequently denote strains by their abbreviated subspecies and name (e.g. *domesticus^WSB^*). **Each mouse from a inbred group was treated as a biological replicate**.

Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice were sampled from an partially-inbred strain of Gough Island mice, approximately 6 generations of brother-sister matting. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC.

@@

The number of cells imaged per individual mouse is based on previous studies [Dumont, Wang]. Images were captured using a Zeiss Axioplan 2 microscope with AxioLab camera and AxioVision software (Zeiss, Cambridge, UK). Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 foci or DMC1 foci using Photoshop.

@@

To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental Figure 1). **During preliminary analysis, cells found to be outliers for MLH1 count had the original image was examined for issues.**

>>>>

We used counts of MLH1 foci per cell to estimate genome-wide recombination rates in `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild-derived inbred strains sampled from three subspecies of house mice ( _M. m. domesticus_, _M. m. musculus_ and _M. m. molossinus_ ) and three other species of Mus ( _M. spretus_, _M. spicilegus_ , and _M. caroli_ ).  Approximately 25 cells were sampled per cell were sampled given time constraints and based on previous studies [@Wang2017island; @dumont2011murid; @murdoch2010]. Mean MLH1 focus counts for `r sum(main.table$Nmice)` mice were quantified from an average of `r mean( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T)` spermatocytes per male (for a total of `r formatC( sum( (Mouse.Table_HQ %>% filter(sex == "male") )$Ncells, na.rm = T), format="d", big.mark=",")` spermatocytes) and `r mean( (Mouse.Table_HQ %>% filter(sex == "female") )$Ncells, na.rm = T)` oocytes per female (for a total of `r formatC(sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells), format="d", big.mark=",")` oocytes) (Table 1). <cite table 1 for MLH2 stats>

@@

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )` early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_) (Table 3). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; p < `r 1/1000000` -- 10-6). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval`). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

@@

We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 4). Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus*, *M. spretus* and _M. caroli_. We subsequently denote strains by their abbreviated subspecies and name (e.g. *domesticus^WSB^*). **Each mouse from a inbred group was treated as a biological replicate**.

Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice were sampled from a partially-inbred strain of Gough Island mice, approximately 6 generations of brother-sister matting. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC.

@@

Images were captured using a Zeiss Axioplan 2 microscope with AxioLab camera and AxioVision software (Zeiss, Cambridge, UK).

**A 
The number of cells imaged per individual mouse is based on previous studies [@Wang2017island; @dumont2011murid; @murdoch2010].

Preprocessing, including cropping, noise reduction, and histogram adjustments, was performed using Photoshop (v13.0). Image file names were anonymized before manual scoring of MLH1 foci or DMC1 foci using Photoshop.

@@

To estimate the number of crossovers across the genome, we counted MLH1 foci. MLH1 foci were counted in cells with intact and complete karyotypes (19 acrocentric bivalents and XY for spermatocytes; 20 acrocentric bivalents for oocytes) and distinct MLH1 foci. A quality score ranging from 1 (best) to 5 (worst) was assigned to each cell based on visual appearance of staining and spread of bivalents. Cells with a score of 5 were excluded from the final analysis. Distributions of MLH1 count per cell were visually inspected for normality (Supplemental Figure 1). When Outliers for MLH1 count were found during preliminary analysis, the original images were inspected and the counts confirmed.
